### 2022-02-12 symposium.tex
Type-Safe Regular Expressions: evaluation

\regexDocumentation
% val date = new Regex("(\\d{4})-(\\d{2})-(\\d{2})")
% "2004-01-20" match
%   case date(year, month, day) =>
%     s"$year was a good year for PLs."

@@

\regexUserLevel
% object Regex:
%   def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
%     new Regex(regex)
%
% class Regex[P] private (val regex: String):
%   val pattern = Pattern.compile(regex)
%   def unapply(s: String): Option[P] =
%     val m = pattern.matcher(s)
%     if (m.matches())
%       val a = Array.tabulate(m.groupCount)(i => m.group(i + 1))
%       Some(transform[P](regex, a))
%     else
%       None

@@

\regexRational
% val rational = Regex("(\\d+)\\.?(\\d+)?")
% "3.1415" match
%   case rational(i, Some(f)) =>
%     val n = i.size + f.size
%     s"This number is $n digits long"

@@

\regexFirstIteration
% import compiletime.ops.string._
% import compiletime.ops.int.+
%
% type Compile[R <: String] =
%   Reverse[Loop[R, 0, Length[R], EmptyTuple]]
%
% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
%   Lo match
%     case Hi => Acc
%     case _  => CharAt[R, Lo] match
%       case "("  => Loop[R, Lo + 1, Hi, String *: Acc]
%       case "\\" => Loop[R, Lo + 2, Hi, Acc]
%       case _ => Loop[R, Lo + 1, Hi, Acc]

@@

\regexIsCapturing
% type IsCapturing[R <: String, At <: Int] <: Boolean =
%   CharAt[R, At] match
%     case "?" => CharAt[R, At + 1] match
%       case "<" => CharAt[R, At + 2] match
%         case "!" => false // zero-width negative lookbehind
%         case _ => true    // named-capturing group
%       case _ => false     // other non-capturing special constructs
%     case _ => true        // unnamed-capturing group

@@

\regexNaiveIsNullable
% type IsNullable[R <: String, At <: Int, Hi <: Int] <: Boolean =
%   CharAt[R, At] match
%     case ")" => IsMarked[R, At + 1, Hi]
%     case _   => IsNullable[R, At + 1, Hi]
%
% type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
%   At match
%     case Hi => false
%     case _ =>
%       CharAt[R, At] match
%         case "?" | "*" => true
%         case "+" => IsMarked[R, At + 1, Hi]
%         case _ => false

@@

\regexIsNullable
% type IsNullable[R <: String, At <: Int, Hi <: Int, Lvl <: Int] <: Boolean =
%   CharAt[R, At] match
%     case ")" => Lvl match
%       case 0 => IsMarked[R, At + 1, Hi]
%       case _ => IsNullable[R, At + 1, Hi, Lvl - 1]
%     case "(" => IsNullable[R, At + 1, Hi, Lvl + 1]
%     case "\\" => IsNullable[R, At + 2, Hi, Lvl]
%     case _    => IsNullable[R, At + 1, Hi, Lvl]

@@

\regexLastIteration
% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =
%   Lo match
%     case Hi => Acc
%     case _  => CharAt[R, Lo] match
%       case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Opt - 1]]
%       case "(" =>
%         Opt match
%           case 0 => IsNullable[R, Lo + 1, Hi, 0] match
%             case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
%             case false => Loop[R, Lo + 1, Hi, Acc, 0]
%           case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Opt + 1]
%       case "\\" => Loop[R, Lo + 2, Hi, Acc, Opt]
%       case _ => Loop[R, Lo + 1, Hi, Acc, Opt]

@@

In Scala, all type parameters are erased, which can lead so some friction when trying to write programs whose execution \emph{depends} on a type, such as in the problem at hand. In this section, we propose two different solutions to that problem.

@@

  \item In \Cref{subsec:implicit-based-extractor-synthesis}, we show...
\end{enumerate}

@@

The first implementation of our runtime component duplicates the type-level definitions presented in the previous section to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
In a nutshell, instead of accumulating |String| and |Option[String]| types, we rewrite the outer loop of our algorithm to accumulate identity functions and eta-expansions of the option constructor:

\regexTermLvlLoop
% val identity: String => Any = { x => assert(x != null); x }
% type Accumulator = List[String => Any]
%
% def loop(r: String, lo: Int, hi: Int, acc: Accumulator, opt: Int): Accumulator =
%   lo match
%     case `hi` => acc
%     case _  => r.charAt(lo) match
%       case ')' => loop(r, lo + 1, hi, acc, 0.max(opt - 1))
%       case '(' =>
%         opt match
%           case 0 => isNullable(r, lo + 1, hi, 0) match
%             case true  => loop(r, lo + 1, hi, Option.apply :: acc, 1)
%             case false => loop(r, lo + 1, hi, identity :: acc, 0)
%           case _ => loop(r, lo + 1, hi, Option.apply :: acc, opt + 1)
%       case '\\' => loop(r, lo + 2, hi, acc, opt)
%       case _ => loop(r, lo + 1, hi, acc, opt)

@@

\regexTransform
% def transform[P](pattern: String, arr: Array[String]): P =
%   val fs = loop(pattern, 0, pattern.length, Nil, 0).reverse
%   val ts = Tuple.fromArray(arr.zip(fs).map { (x, f) => f(x) })
%   assert(arr.size == fs.size, "Unexpected number of capturing groups")
%   ts.asInstanceOf[P]

@@

The first alternative that comes to mind is perhaps to turn ourselves to a dependently typed language that supports type- and term-level polymorphism.
In a language with shared term and type syntax, we could write our analysis generically and derive two programs from it, one for each level.

In principle, Scala~3's type inference should be able to solve half of the problem with its ability to correlate match types and match expressions.
In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.
In our case, this mechanism should allow us to get rid of the unsafe cast but does not address the problem of code duplication.
In practice, at the time of writing, this solution does not play well with predefined types from the standard library, because the compiler lacks the knowledge necessary to correlate those types with their term-level counterpart.
For instance, the compiler does not recognize |CharAt["foo",0]| as a valid type for |"foo".chatAt(0)|.

In the long run, it would be interesting to investigate extending Scala with a "precise" mode of type inference, that would automatically infer match types and singleton types, whenever possible.
In \Cref{chap:generalizing-singleton-types}, we demonstrated the feasibility of this approach by generalizing Scala's singleton types.
A promising avenue would be to revisit that line of work through the lens of type inference.

@@

The second implementation of our runtime component takes a completely different approach.
Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
That synthesis is type-directed: it generates a single-purpose program that sanitized the output of the regex engine for a particular regular expression.

Our implementation consists of a type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements.
The |Sanitizer| type class defines a single method that mutates an array in place to warp its nullable elements in options:

\regexSanitizerTypeClass
% abstract class Sanitizer[T](val i: Int):
%   def mutate(arr: Array[Any]): Unit
%
% object Sanitizer:
%   implicit val basecase: Sanitizer[EmptyTuple] =
%     new Sanitizer[EmptyTuple](0):
%       def mutate(arr: Array[Any]): Unit = ()
%
%   implicit def stringcase[T <: Tuple]
%     (implicit ev: Sanitizer[T]): Sanitizer[String *: T] =
%       new Sanitizer[String *: T](ev.i + 1):
%         def mutate(arr: Array[Any]): Unit =
%           assert(arr(arr.size - i) != null)
%           ev.mutate(arr)
%
%   implicit def optioncase[T <: Tuple]
%     (implicit ev: Sanitizer[T]): Sanitizer[Option[String] *: T] =
%       new Sanitizer[Option[String] *: T](ev.i + 1):
%         def mutate(arr: Array[Any]): Unit =
%           arr(arr.size - i) = Option(arr(arr.size - i))
%           ev.mutate(arr)

From a design standpoint, this solution is appealing as it leads to no duplicated code and no duplicated computations: we perform our capture group analysis at the type-level, once and for all, and synthesize code accordingly.
This combination of type-level programming and implicit-based code synthesis is reminiscent of the compile-time variant of multi-stage metaprogramming \citep{stucki2018practical}.
In those terms, our approach corresponds to a multi-staged program whose stage 0 happens entirely within Scala's type checker.

@@

To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
We select the subset of those tests that are positive (they match successfully) and exercises capturing groups, totaling 183 test cases, which we then collated to form a single benchmark.
We use this benchmark to evaluate the performance of our library in terms of compilation times and execution times.

\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code duplicated runtime (Dup) and with its implicit-based runtime (Impl). Lower is better.}

In \Cref{fig:regexBars}, we show the results of our experiments, which compare Scala's standard regular expression library (Std) against two flavors of our library, one using our code duplicated runtime (Dup), presented in \Cref{subsec:we-ain-t-need-no-dependent-types}, and the other using our implicit-based runtime (Impl), presented in \Cref{subsec:implicit-based-extractor-synthesis}.

XXX

We obtained these results by averaging the measurements obtained over 2 hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
With these settings, we obtains margins of errors below \textpm1\%, with 99.9\% confidence (omitted from \Cref{fig:regexBars}).

\section{Related Work}

>>>>

\regexDocumentation

@@

\regexUserLevel

@@

\regexRational

@@

\regexFirstIteration

@@

\regexIsCapturing

@@

\regexNaiveIsNullable

@@

\regexIsNullable

@@

\regexLastIteration

@@

In Scala, all type parameters are erased, which leads to some friction when wanting to write programs whose execution \emph{depends} types, such as in the problem at hand. In this section, we propose two different solutions to that problem.

@@

  \item In \Cref{subsec:implicit-based-extractor-synthesis}, we show how to use implicits to perform type-directed code synthesis: we use the output of our type-level analysis as an input of implicit search to generate a type-specialized capture group sanitizer.
\end{enumerate}

@@

The first implementation of our runtime component duplicates the type-level definitions presented in the previous section to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
In a nutshell, instead of accumulating |String| and |Option[String]| in a type, we accumulate identity functions and eta-expansions of the option constructor in a list:

\regexTermLvlLoop

@@

\regexTransform

@@

The first alternative that comes to mind is perhaps to turn ourselves to a dependently typed language with support for type- and term-level polymorphism.
In a language with shared term and type syntax, we could write our analysis generically, and derive two programs from it, one for each level.

In principle, Scala~3's type inference should be able to solve half of that problem with its ability to correlate match types and match expressions.
In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.
In our case, this mechanism should allow us to get rid of the unsafe cast but does not address the problem of code duplication.
In practice, at the time of writing, this solution does not play well with predefined types from the standard library: the compiler lacks the knowledge necessary to correlate predefined types with their term-level counterpart.
For instance, the compiler does not recognize |CharAt["foo", 0]| as a valid type for |"foo".chatAt(0)|.

In the long run, it would be interesting to investigate extending Scala with a "precise" mode of type inference, that would automatically infer match types and singleton types, whenever possible.
In \Cref{chap:generalizing-singleton-types}, we demonstrated the feasibility of this approach by generalizing Scala's singleton types.
A promising avenue would be to revisit that line of work with the less ambitious goal of improving Scala's type inference.

@@

The second version of our runtime component takes a completely different approach.
Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
That synthesis is type-directed: it generates a single-purpose program that sanitized the output of the regex engine by following the shape of the tuple type computed at the type level.

Our implementation consists of one type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements (here, |T| is the results of our type-level analysis).
The |Sanitizer| type class defines a single method that mutates an array in place to warp nullable elements in options:

\regexSanitizerTypeClass

From a design standpoint, this solution is appealing as it leads to no duplicated code or computation: we perform our capture group analysis at the type-level, once and for all, and synthesize code accordingly.
This combination of type-level programming and implicit-based code synthesis is reminiscent of multi-stage metaprogramming, in its compile-time variant \citep{stucki2018practical}.
In those terms, our approach corresponds to a multi-staged program whose stage 0 happens entirely within Scala's type checker.

@@

To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
We select the subset of those tests that are positive (they match successfully) and exercises capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
We use these benchmarks to evaluate the performance of our library in terms of compilation times and execution times.

In \Cref{fig:regexBars}, we show the results of our experiments, which compare Scala's standard regular expression library (Std) against two flavors of our library, one using our code-duplicated runtime (Dup), presented in \Cref{subsec:we-ain-t-need-no-dependent-types}, and the other using our implicit-based runtime (Impl), presented in \Cref{subsec:implicit-based-extractor-synthesis}.

\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code-duplicated runtime (Dup) and with its implicit-based runtime (Impl). Lower is better.}

We obtained these results by averaging the measurements obtained over 2 hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
With these settings, we obtains margins of errors below \textpm1\%, with 99.9\% confidence (omitted in \Cref{fig:regexBars}).
We run our benchmarks on the latest version of Scala~3 (3.1.2 at the time of writing).

\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline, which corresponds to an average of +0.3~ms per regex.
This corresponds to the cost of match type reduction.
The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost implicit resolution.
Both variants are thus relatively cheap which validates their utility for practical uses.
When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of the type checker \citep[2.11.3]{petrashko2017design}.

\paragraph{Execution Times} At runtime, the code-duplicated variant adds 51~\textmu s over the baseline.
Although this difference is three orders of magnitudes smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis.
The implicit-based variant gets rid of this cost by synthesizing a type-specialized runtime (at the cost of increased compilation timed).
To our surprise, our implicit-based runtime outperforms Scala's standard library implementation.
We suspect that this difference is due to the poor interaction between the JVM's just-in-time compiler and the code generated for variadic extractors (|unapplySeq|, used by the Std implementation).
Our implicit-based implementation appears to be more prone to inlining and partial evaluation since it does not involve generic sequences, which could explain the slight grain in performance.

\bigskip

In view of these results, we decided to use our implicit-based runtime in the published version of our library.
\section{Related Work}
